---
title: "individual-assignment5"
author: "sofia favela"
format: pdf
date: 5/8/2026
---

# Set Up 

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(janitor)
# file organization
library(here)
# wrapping long axis labels
library(scales)
# reading in .xlsx files
library(readxl)
# visualizing missing data
library(naniar)
# arranging multiple plots
library(patchwork)
# converting to snake case
library(snakecase)
```

## Data
```{r}
# taxonomic information for each taxon ID
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrate count data
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects")

# site metadata
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality measurements
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+"))
```

# a. Planning the figure 

- x axis: `site_full` ( site name)
- y axis: log + 1 transformed average ostracod abundance per liter 
- geometry to represent average abundance/L: 'geom_jitter() to plot every individual survey observation as points
- geometry to represent medians: stat_summary(fun = "median") overlays colored points to show medians per site 

# b. Wragling Data
 - filter aq_ins_clean to only include ostracods
 - log +1 transform average abundances 

```{r}
#| label: wrangling-aq-ins_clean_data_frame

# creating new object from aq_ins_clean
ostracods <- aq_ins_clean |> 
  # filter to only include the taxon titled ostracod
  filter(taxon == "ostracod") |> 
  # creates new column called log_ave_lit that contains the log + 1 transformed average abundance per liter 
  mutate(log_ave_lit = log(ave_lit + 1))

slice_sample(ostracods, n = 10)
```
 
# c. Coding the figure
```{r}






```


